The Evolution of Phenotypically Invariant Genetic Networks

### 2021-04-27 review-responses.tex

\begin{point}{}
    Your analysis of the fitness loss in F1 and F2 hybrids rests on a local
Taylor expansion of the fitness landscape. This is adequate given that the
underlying linear network is also only locally valid (see above). However, you
then apply this to a discussion of hybrid incompatibility (Haldane's rule etc).
We are thus interested in \emph{very large} fitness costs for hybrids. This does not
seems to be compatible: to get hybrid incompatibility or sterility, you need to
have epsilon sufficiently large (below equation 5). But then $\epsilon^2$ is no
longer smaller than $\epsilon$. In other words: it seems to me that for the
discussion of hybrid incompatibilities you apply your model of the first part
beyond the local range where it is valid.
\end{point}

\reply{
    This is a good point, but tricky since it's not clear what ``small'' is here.
    A fitness difference of 0.2 is quite strong selection, for instance,
    although it is a fairly small number in some ways.
    Furthermore, the same picture holds in the opposing cases of a single locus
    or many loci, illustrated in Figure~\ref{fig:conceptual_fig}.
    In any case, we now rely only on the conceptual idea rather than the numerics of this estimate.
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
The present paper is divided into two parts. In the first part, the authors propose a framework
in which distinct genetic architectures can produce the same phenotype. (In the fitness landscape
terminology, this would correspond to the existence of an evolutionary ridge in a fitness landscape.)
The starting point is a nice analogy with linear systems theory. The authors highlight the fact
that in general, two distinct linear (differential) systems can respond identically for any input; i.e.,
different systems (genotypes) can always produce the same output (phenotype) given the same input
(environment). Further, the set of equivalent systems can be nicely characterized through Kalman
decomposition, thus providing a nice characterization of level sets in the underlying fitness landscape.
All along the first part of the paper, this analogy is well illustrated using a simple (yet quite nice)
example of oscillating gene transcription. I especially like Fig. 5 where it is shown that even if $F_1$
hybrids have a phenotypic response close to their parents, $F_2$ hybrids can behave in a drastically
different manner.

@@

In the second part of the paper, motivated by the previous results, the authors explore a general
quantitative genetics model in which populations can drift stochastically near a set of equivalent
and optimal systems. Since the optimal set (or evolutionary ridge) is not closed under averaging or
recombination, two isolated populations can drift apart and accumulate enough genetic differences
so that they do not produce any viable offspring. Using some heuristics, several expressions are
derived to quantify the accumulation of genetic incompatibilities.

Overall, I think the paper is well written. The framework developed in the first part of the paper
is very interesting and that the analogy with system theory is quite enlightening. I am somehow
less convinced by the second part. First, I find the arguments a bit sketchy (see below) and not so
easy to follow. Secondly, it is not entirely clear to me what is the main contribution of this part
compared to previous works. It seems to me that the main result is somehow contains in the fact
that the variance (or ``segregation variance'') of an $F_2$ population is given by
$$\sigma^2_S + 4 \omega \sigma^2_N T / N_e$$
which was already derived Slatkin and Lande according to the authors (except for the explicit
expression of $\omega$, but again I am a little bit confused by the arguments derived in the appendix).

In summary, I think this paper could be a nice contribution to Evolution. However, I am also
convinced that the second part of the paper would require more work, or more explicit reference
to previous works (equation, section etc.) if the authors do not want to re-derive already existing
formula.
\end{quote}

Thanks for the encouraging words.
As discssed above, we've separated out the second (less convincing) part,
and the manuscript now only covers the first part.

\begin{point}{end of p6:} 
    ``we assume ... $((A + A' )/2, (B + B' )/2, (C + C' )/2)$'' . It is claimed \revreffull{2}{1}
    that two kryptotypes need not have the same dimension. In this
    case, the previous sum does not make sense, right?
\end{point}

@@

\reply{
    Good point; added that caveat. \revref
}

\begin{point}{p7: fig 4.}
    ``The distribution of $F_2$ ... homozygotes''. I do not really grasp the meaning of this
sentence. More importantly, I find the purple cloud of dots ($F_2$ population) quite confusing.
It seems to me that the purple populations should be a cloud of points concentrated around
a point sitting on the red dotted line, i.e., a Gaussian distribution with the mean at the
average of the two parental populations. According to the authors, the distribution is
bimodal with a peak below the optimal set (as displayed on the figure) and another distinct
peak sitting on the other side of the optimal. In fact, if I understand the computations of
the second part correctly, the red and purple means should coincide, but the red variance
should simply be greater.
\end{point}

\reply{
    We have improved Figure~\ref{fig:conceptual_fig} and made the accompanying discussion
    more concrete to hopefully better demonstrate the point.
}

@@

\reply{
    This part has been removed.
}

\begin{point}{Fig 7.}
    I have one important issue with this figure (and the assumptions of the underly-
ing quantitative genetic model). If one wants to be consistent with the assumption that
parental populations drift along the evolutionary ridge, I think one would need to assume
that selection is strong enough to constraint the mean of the population on the surface. This
would presumably require that $\sigma S /\gamma \ll 1$. First, I think this assumption (or something
alternative to that) should be made explicit in the text. Secondly, it seems to me that this
assumption is not satisfied for panel A and C: under the range of parameters proposed by
the authors, the parental populations could easily drift away from the optimal set, and in
particular, the heuristics derived in the main text would not be satisfied.
\end{point}

\reply{
    This part has been removed.
}

\begin{point}{}
    Finally, it would be worth mentioning several old and recent works relating genetic drift
    to speciation: \citet{yamaguchi2013first}, several articles by Gavrilets et al. (I think several
    citations are missing here), and \citet{pina2019does}.
\end{point}

### 2021-04-27 kalman_walk_paper.tex
rev 2 edits

In this paper we use results on mathematical nonidentifiability from
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}.
% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.

>>>>

In this paper we use results on mathematical nonidentifiability from
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  \revpoint{2}{12}
% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.
